Polynomial Time and Dependent Types

### 2023-11-14 paper/paper.tex
[ paper ] final edits

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Polynomial Time and Dependent Types}
% \subtitle{Extended Version}

@@

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
To construct a $\conZero$, we must have a $\Diamond$ to pay for it,
and likewise, to construct a $\conSucc$ we must pay a $\Diamond$. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \emph{Cons-free}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \emph{Cons-free} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an additional binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to \citet{AehligS02},
illustrates how this leads to all polytime computations. As above, we
assume a polynomial $p(n)$ and a step function
$f : \mathit{St} \multimap \mathit{St}$ that needs to be iterated
$p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \emph{Cons-free} system, this iterator does not raise
the degree of the nested iterator directly. Rather, the iterator $I_k$
on the input $n$ performs $\binom{n}{k}$ iterations. As observed by
Aelhig and Schwictenberg, this is sufficient because the binomials
form a basis for the vector space of all polynomials.

@@

In this section, we give an overview of the rules of QTT. The full
rules, including all equality rules, are presented in
%\autoref{sec:rules}.
the online appendix \cite{atkey2023polydep-online}.

@@

The fact that we can characterise polytime decision problems is
perhaps to be expected from a system designed to capture polynomial
time realisable programs. However, we can go further to capture the
complexity classes of Non-deterministic Polynomial time (NP) and
Probabilistic Polynomial time (PP), both of which are based on
polytime. We do this by augmenting our polytime functions with
additional power in the form of computation effects.

@@

To capture the complexity class NP, we use polynomial time programs
augmented with non-determinism, as one might expect. We will not need
to reason about equality of these non-deterministic programs, so we
can represent non-deterministic choices as \emph{additive} binary
trees. We suppose a non-iterable datatype defined like so:
\begin{displaymath}
  \begin{array}{l}
    \textbf{data}\,\mathrm{ND}\,(A : \mathsf{U}) : \mathsf{U}\,\textbf{where} \\
    \quad \mathsf{return} : A \to \mathrm{ND}\,A \\
    \quad \mathsf{choice} : (\BoolTy \to \mathrm{ND}\,A) \to \mathrm{ND}\,A
  \end{array}
\end{displaymath}
The crucial point here is that the subtrees are represented as a
function $\BoolTy \to \mathrm{ND}\,A$. By the typing rules of QTT,
this means that the two branches of this function can share resources
(see the encoding of the additive product types by
\citet{atkey18qtt}). Thus, each branch of this tree is polynomial
time, not the whole tree itself.

@@

Probabilistic Polynomial time has previously been considered in the
setting of implicit computational complexity by
\citet{dallago_et_al:LIPIcs.MFCS.2021.35} and \citet{LagoT15}. In both
cases, they must build probabilistic choice into the language, and
have difficulty in directly capturing the class BPP due to its
semantic nature \cite{AroraBarak09}, where the correctness of
implementation is probabilistic. With a dependently-typed host
language, adding probabilistic choice as an effect and capturing the
semantic constraint of BPP is straightforward.

@@

In this section and the next, we establish the polytime soundness of
our extensions of QTT by adapting a realisability method due to
\citet{dallago11realisability}. This approach is based on a three way
coupling between abstract mathematical elements (the \emph{what}),
values from a machine model (the \emph{how}), and resource potentials
(the \emph{fuel}). Each type in the system is defined as a three way
relation between these elements. The set of abstract elements depends
on the type being interpreted (e.g., types of natural numbers will be
defined in terms of the set $\mathbb{N}$). The machine model is fixed
across all types. We describe the particular machine model we use for
this paper in \autoref{sec:machine-model}. Potentials are arranged
into \emph{resource monoids} that we define in
\autoref{sec:resource-monoids}. Unlike \citet{dallago11realisability},
we explicitly construct realisers for inductive datatypes (both
iterable and non-iterable) instead of relying on second-order
polymorphic encodings and special $\oc$-style modalities. These
explicit constructions are essential to construct a model of dependent
types.

\paragraph{Agda Formalisation} The key soundness results in this
section have been formalised in the Agda proof assistant
\cite{norell2008dependently}. The Agda formalisation can be found in
the supplementary material. After each definition and result we
provide a pointer to the Agda modules where the corresponding
formalisation can be found, and note interesting features of the
mechanisation.

@@

It is perhaps easier to see how the differencing operator works in the
special case of the difference $\MaxPoly((m,p),(0,0)) = p(m)$. I.e.,
if we have code that contains data of size $m$ and a program with
complexity $p$, then running the combination with no expectation of
remaining potential yields $p(m)$ available steps. The $\MaxPoly$ and
$\PlusPoly$ resource monoids will be used for the \emph{Cons-free} and
LFPL-style systems respectively, as we explain in
\autoref{sec:realising-iteration} and show how these resource monoids
yield the required polytime bounds on programs.

@@

We now define an indexed linear poset $L$ of realisers over $\Set$
that ties together our ``mathematical'' model of types in $\Set$ with
our machine model and resource monoid. This construction is a
reformulation of \citet{dallago11realisability}'s realisability models
suitable for modelling dependent types. For a set $A$, the carrier of
$L(A)$ is the set of ternary relations
$X \subseteq A \times M \times \ValSet$ and we define the ordering
$X \leq Y$ to hold iff there exists a realising expression
$E \in \ExpSet$ and potential $\gamma \in M_0$ such that for all
$a \in A$, $\alpha \in M$ and $v \in \ValSet$ with
$(a,\alpha,v) \in X$, we have that there exists a result
$v' \in \ValSet$, step count $k \in \mathbb{N}$ and result potential
$\beta \in M$ with:
\begin{enumerate}
\item $E, v \Downarrow_k v'$ (evaluation successfully completes in $k$ steps);
\item $(a, \beta, v') \in Y$ (the result is well-resourced and
  satisfies $Y$); and
\item $k \leq M(\alpha \rplus \gamma, \beta)$ (the step count is within the
  difference between the initial potential and the result potential).
\end{enumerate}
Note that the definition of realisablity is uniform in the element $a$
-- the realising expression $E$ and the potential $\gamma$ must work
for all $a$ -- thus the implementation and complexity measure of the
transition being modelled cannot depend on what the input is. Put in
implementation terms, the input $a$ is not present at
runtime. Moreover note that the potential $\gamma$ attached to the
expression $E$ must come from the sub-monoid $M_0$, indicating that is
intended to be data-free, while the potential $\alpha$ for the input
is from the full monoid $M$, so it can contain data and functions.

@@

Before that, we show how to define realisers for the representative
examples of non-iterable types from \autoref{sec:linear-case-analysis}
and \autoref{sec:noniter-qtt}. Booleans are the simplest case, with
only two cases and no chance of iteration. Lists are more complex; we
can have non-iterable lists containing iterable data.

@@

The models constructed in the previous section only allow for
constant-time programs to be realised. To interpret the iterators of
the \emph{Cons-free} and LFPL-style systems, we need to use the
$\MaxPoly$ and $\PlusPoly$ resource monoids. We do this in this
section, where first we establish some operations that will be useful
to see how they capture the nesting of iterations inherent to polytime
computation.

@@

  \mathrm{natValue}(1 + n) = (\synFalse, \mathrm{natValue}(n))
\end{mathpar}
This representation uses a tagged pair approach similar to our
representation of lists in \autoref{sec:noniter-model}. Using this, we
can define what it means for a natural number to be realisable via
$\mathrm{Nat} \in L(\mathbb{N})$:
\begin{displaymath}
  \mathrm{Nat} = \{(n, \alpha, \mathrm{natValue}(n)) \mid n \in \mathbb{N}, 0 \leq M(\alpha, \mathit{size}(n+1)) \}
\end{displaymath}
So a natural number $n$ is realised by the value
$\mathrm{natValue}(n)$ as long as we have at least
$\mathit{size}(n+1)$ potential (we add one to make the LFPL soundness
proof easier).  This gives us the ability to represent natural numbers
as a type in QTT, but in order to iterate (and construct in the case
of LFPL), we need to construct specific realisers for the
\emph{Cons-free} and LFPL systems.

@@

The \emph{Cons-free} system uses the $\MaxPoly$ resource monoid, with
the distinguished sub-monoid being those elements that are $0$ in the
size component. We enumerate the features of the \emph{Cons-free}
system and justify their realisability with the $\MaxPoly$ resource
monoid:
\begin{enumerate}
\item Duplication of natural numbers by $\dupNat(M)$ is realisable by
  the expression $(0,0)$, which creates a pair by copying the input
  variable twice. By the cost semantics in
  \autoref{sec:machine-model}, this takes $1$ step of computation (we
  assume that it is actually implemented via some pointer copy).  The
  resource accounting for this realiser works because the size
  component required for the output is the maximum of the size
  components of the two elements, and since $n \sqcup n = n$, we have
  enough resources to fulfil this.
\item Construction of natural numbers is not realisable. In a putative
  $\conSucc$ rule, we would need to get an additional unit of size
  resource from nowhere.
\item Iteration is realised by constructing a realising expression in
  the expression language from the given expressions for the zero and
  successor cases that uses the in-built recursion of the
  language. The proof that resources are correctly accounted for is
  carried out by induction on the natural being iterated over. For
  $n$, we require potential
  $\mathit{scale}(n,\mathit{acct}(4) \rplus \gamma_{\mathit{succ}})
  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$, where
  $\gamma_{\mathit{succ}}$ and $\gamma_{\mathit{zero}}$ are the
  potentials required by the successor and zero cases respectively. By
  Property (2) of iteration resource monoids, above, we know that
  $\mathit{raise}(\mathit{acct}(4) \rplus \gamma_{\mathit{succ}})
  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$ always
  dominates this requirement when paired with the potential
  $\mathit{size}(n)$ from the input. Therefore, this latter
  expression, plus some administrative set up costs, is the required
  potential for the whole iterator.
\end{enumerate}
Together, we have a soundness result for the \emph{Cons-free} system,
that ensures that every term in the $\sigma = 1$ fragment is
realisable by a \emph{correct} program that terminates in polynomial
time for all inputs:
\begin{theorem}[Soundness for the Cons-free System]
  \label{thm:cons-free-soundness}
  If we have a term $n : \tyNat \vdash M \stackrel1: T(n)$ then there
  exists a realising expression $E$ and polynomial $p$ such that for
  all $n \in \mathbb{N}$, there exists $v \in \ValSet$ and
  $k \in \mathbb{N}$ such that
  $E, [\mathrm{natValue}(n)] \Downarrow_k v$, $k \leq p(n)$ and $v$ is
  a realising value for
  $\llbracket M \rrbracket(n) \in \llbracket T \rrbracket(n)$.
\end{theorem}

\paragraph{Agda Formalisation} The realisability of the
\emph{Cons-free} system iterator and the soundness property of the
whole system are formalised in the Agda modules \AgdaModule{ConsFree}
and \AgdaModule{ConsFree.Iterator}. The soundness theorem is a
combination of this and the QTT model sketched in
\autoref{sec:qtt-models}.

@@

The LFPL system uses the $\PlusPoly$ resource monoid, with the
distinguished sub-monoid again being those elements that are $0$ in
the size component. With this resource monoid, the capabilities
offered at the QTT level are altered:
\begin{enumerate}
\item We can no longer duplicate natural numbers, because
  $\tyNat \otimes \tyNat$ requires twice as much size resource as
  $\tyNat$, due to the combining operation on size potentials being
  addition.
\item We define the realisability specification for diamonds
  $\Diamond \in L(1)$ to be
  $\Diamond = \{ (*, \alpha, *) \mid 0 \leq M(\alpha,
  \mathit{size}(1)) \}$.  Thus, a diamond represents at least one unit
  of size resource, matching the intuitive explanation given in
  \autoref{sec:lfpl-intro}.
\item With this definition of realisability for $\Diamond$s, it is
  possible to realise the $\conZero$ and $\conSucc$ constructors for
  natural numbers. By the additive combination of size resources we
  get $1$ from the diamond and $n+1$ from the predecessor to total
  $n+2$ for a new number. Note that, even if we add a $\Diamond$ type
  to the \emph{Cons-free} system, it would still not be possible to
  realise the constructors, because we would only have
  $1 \sqcup (n+1) = n+1$ size resource for the output.
\item The construction of the recursor follows a very similar proof to
  the realisability of \emph{Cons-free} iterator, up to some
  additional work to make sure that the dummy $*$ values representing
  the diamond components end up in the right places. This additional
  work is revealed in the required potential for the LFPL iterator
  being
  $\mathit{raise}(\mathit{acct}(8) \rplus \gamma_{\mathit{succ}})
  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$, so
  slightly higher in the successor case.
\end{enumerate}

Soundness for the LFPL system is similar to the \emph{Cons-free}
system, except for a $+1$ to the input to the polynomial, to account
for the fact that we cost one size unit for the $\conZero$
constructor.
\begin{theorem}[Soundness for the LFPL-style System]
  \label{thm:lfpl-soundness}
  If we have a term $n : \tyNat \vdash M \stackrel1: T(n)$ then there
  exists a realising expression $E$ and polynomial $p$ such that for
  all $n \in \mathbb{N}$, there exists $v \in \ValSet$ and
  $k \in \mathbb{N}$ such that
  $E, [\mathrm{natValue}(n)] \Downarrow_k v$, $k \leq p(n+1)$ and $v$
  is a realising value for
  $\llbracket M \rrbracket(n) \in \llbracket T \rrbracket(n)$.
\end{theorem}

The proofs of well-accounted realisability for the LFPL iterator, and
the \emph{Cons-free} iterator, could be adapted to any other
inductively defined type that is finitely branching. This is not
immediately necessary, as evidenced by the construction of other
datatypes in \autoref{sec:prog-datatypes}. Nevertheless, native tree
type where the iterability is proportional to the total number of
nodes would be useful.

@@

We have presented two extensions of Quantitative Type Theory that
soundly and completely capture polynomial time. We have seen that this
allows for an expressive combination of verification and complexity
constrained computation. We now discuss some related work in implicit
computational complexity and linear dependent types, and take a look
at where the combination of resource conciousness and dependent types
could take us.

@@

\paragraph{Implicit Computational Complexity with Linear Types}
Implicit Computational Complexity \cite{Lago11} is a vast field, so we
only survey closely related works. We have already mentioned the
Bounded Linear Logic \cite{bll92}, Soft Affine Logic \cite{Lafont04},
Light Linear Logic \cite{lll98} and LFPL \cite{hofmann99lfpl} systems,
which all use linear typing to implicitly capture polynomial
time. \citet{Jones01} characterises polynomial time using first-order
functional programs without constructors. Thus it shares a method with
our \emph{Cons-free} system, but we use linear typing to permit
controlled use of higher-order functions. Other approaches to
polynomial time use stratification or information flow tracking to
ensure that the outputs of iteration may not be used unrestrictedly to
drive further iteration. For example, \cite{BellantoniC92} and
\cite{HainryP23}. Below polynomial time, systems have be devised to
capture LOGSPACE \cite{LagoS16}. Above polynomial time, systems such
as Elementary Affine Logic (EAL) capture all Elementary-time functions
\cite{CoppolaM01}.

@@

\paragraph{Other Complexity Classes, and Explicit Resource Tracking}
We have been able to characterise the classes NP and BPP in terms of
our underlying characterisation of P (\autoref{sec:classes}). It seems
straighforward to extend this to related classes like coNP, RP,
etc. It also seems feasible to adapt the techniques presented here to
other complexity classes such as LOGSPACE and ELEMENTARY, given the
simply typed linear systems mentioned above. Complexity classes based
on circuits may be more challenging, but we do now have a way to
characterise circuits that are generatable in polynomial
time.

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Polynomial Time and Dependent Types}
%\subtitle{Extended Version}

@@

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
To construct a $\conZero$, we must have a $\Diamond$ to pay for it,
and likewise, to construct a $\conSucc$ we must pay a $\Diamond$. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \ConsFree{}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \ConsFree{} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an additional binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to \citet{AehligS02},
illustrates how this leads to all polytime computations. As above, we
assume a polynomial $p(n)$ and a step function
$f : \mathit{St} \multimap \mathit{St}$ that needs to be iterated
$p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \ConsFree{} system, this iterator does not raise
the degree of the nested iterator directly. Rather, the iterator $I_k$
on the input $n$ performs $\binom{n}{k}$ iterations. As observed by
Aehlig and Schwictenberg, this is sufficient because the binomials
form a basis for the vector space of all polynomials.

@@

In this section, we give an overview of the rules of QTT. The full
rules, including all equality rules, are presented in
%\autoref{sec:rules}.
the exteneded version of this paper \cite{atkey2023polydep-online}.

@@

The fact that we can characterise polytime decision problems is
perhaps to be expected from a system designed to capture polynomial
time realisable programs. However, we can go further to capture the
complexity classes of Non-deterministic Polynomial time (NP) and
Probabilistic Polynomial time (PP), both of which are based on
polytime. We do this by augmenting our polytime functions with
additional power in the form of computational effects.

@@

To capture the complexity class NP, we use polynomial time programs
augmented with non-determinism, as one might expect. We will not need
to reason about equality of these non-deterministic programs, so we
can represent non-deterministic choices as binary trees. We suppose a
non-iterable datatype defined like so:
\begin{displaymath}
  \begin{array}{l}
    \textbf{data}\,\mathrm{ND}\,(A : \mathsf{U}) : \mathsf{U}\,\textbf{where} \\
    \quad \mathsf{return} : A \to \mathrm{ND}\,A \\
    \quad \mathsf{choice} : (\BoolTy \to \mathrm{ND}\,A) \to \mathrm{ND}\,A
  \end{array}
\end{displaymath}
The crucial point here is that the subtrees are represented as a
function $\BoolTy \to \mathrm{ND}\,A$. By the typing rules of QTT,
this means that the two branches of this function can share resources
(see the encoding of the additive product types by
\citet{atkey18qtt}). Thus, each branch of this tree can be explored in
polynomial time, but not the whole tree itself.

@@

Probabilistic Polynomial time has previously been considered in the
setting of implicit computational complexity by
\citet{dallago_et_al:LIPIcs.MFCS.2021.35} and \citet{LagoT15}. In both
cases, they must build probabilistic choice into the language, and
have difficulty in directly capturing the class BPP due to its
semantic nature, where the correctness of implementation is
probabilistic. With a dependently-typed host language, adding
probabilistic choice as an effect and capturing the semantic
constraint of BPP is straightforward.

@@

In this section and the next, we establish the polytime soundness of
our extensions of QTT by adapting a realisability method due to
\citet{dallago11realisability}. This approach is based on a three way
coupling between abstract mathematical elements (the \emph{what}),
values from a machine model (the \emph{how}), and resource potentials
(the \emph{fuel}). Each type in the system is defined as a three way
relation between these elements. The set of abstract elements depends
on the type being interpreted (e.g., types of natural numbers will be
defined in terms of the set $\mathbb{N}$). The machine model is fixed
across all types. We describe the particular machine model we use for
this paper in \autoref{sec:machine-model}. Potentials are arranged
into \emph{resource monoids} that we define in
\autoref{sec:resource-monoids}. Unlike \citet{dallago11realisability},
we explicitly construct realisers for inductive datatypes (both
iterable and non-iterable) instead of relying on second-order
polymorphic encodings and special $\oc$-style modalities. These
explicit constructions are essential to construct models of our systems.

\paragraph{Agda Formalisation} The key soundness results in this
section have been formalised in the Agda proof assistant
\cite{norell2008dependently}. The Agda formalisation can be found in
the associated artefact \cite{atkey2023polydep-artefact}. After each
definition and result we provide a pointer to the Agda modules where
the corresponding formalisation can be found, and note interesting
features of the mechanisation.

@@

It is perhaps easier to see how the differencing operator works in the
special case of the difference $\MaxPoly((m,p),(0,0)) = p(m)$. I.e.,
if we have code that contains data of size $m$ and a program with
complexity $p$, then running the combination with no expectation of
remaining potential yields $p(m)$ available steps. The $\MaxPoly$ and
$\PlusPoly$ resource monoids will be used for the \ConsFree{} and
LFPL-style systems respectively, as we explain in
\autoref{sec:realising-iteration} and show how these resource monoids
yield the required polytime bounds on programs.

@@

We now define an indexed linear poset $L$ of realisers over $\Set$
that ties together our ``mathematical'' model of types in $\Set$ with
our machine model and resource monoid. This construction is a
reformulation of \citet{dallago11realisability}'s realisability models
to make it suitable for dependent types. For a set $A$, the carrier of
$L(A)$ is the set of ternary relations
$X \subseteq A \times M \times \ValSet$ and we define the ordering
$X \leq Y$ to hold iff there exists a realising expression
$E \in \ExpSet$ and potential $\gamma \in M_0$ such that for all
$a \in A$, $\alpha \in M$ and $v \in \ValSet$ with
$(a,\alpha,v) \in X$, we have that there exists a result
$v' \in \ValSet$, step count $k \in \mathbb{N}$ and result potential
$\beta \in M$ with:
\begin{enumerate}
\item $E, v \Downarrow_k v'$ (evaluation successfully completes in $k$ steps);
\item $(a, \beta, v') \in Y$ (the result is well-resourced and
  satisfies $Y$); and
\item $k \leq M(\alpha \rplus \gamma, \beta)$ (the step count is within the
  difference between the initial potential and the result potential).
\end{enumerate}
Note that the definition of realisablity is uniform in the element $a$
-- the realising expression $E$ and the potential $\gamma$ must work
for all $a$ -- thus the implementation and complexity measure of the
transition being modelled cannot depend on what the input is. Put in
implementation terms, the input $a$ is not present at
runtime. Moreover note that the potential $\gamma$ attached to the
expression $E$ must come from the sub-monoid $M_0$, indicating that is
intended to be data-free, while the potential $\alpha$ for the input
is from the full monoid $M$, so it can contain data and functions.

@@

Before that, we show how to define realisers for the representative
examples of non-iterable types from \autoref{sec:linear-case-analysis}
and \autoref{sec:noniter-qtt}. Booleans are the simplest case, with
only two cases and no chance of iteration. Lists are more complex: we
can have non-iterable lists containing iterable data.

@@

The models constructed in the previous section only allow for
constant-time programs to be realised. To interpret the iterators of
the \ConsFree{} and LFPL-style systems, we need to use the
$\MaxPoly$ and $\PlusPoly$ resource monoids. We do this in this
section, where first we establish some operations that will be useful
to see how they capture the nesting of iterations inherent to polytime
computation.

@@

  \mathrm{natValue}(1 + n) = (\synFalse, \mathrm{natValue}(n))
\end{mathpar}
This representation uses a tagged pair approach similar to our
representation of lists in \autoref{sec:noniter-model}. Using this, we
can define what it means for a natural number to be realisable via
$\mathrm{Nat} \in L(\mathbb{N})$:
\begin{displaymath}
  \mathrm{Nat} = \{(n, \alpha, \mathrm{natValue}(n)) \mid n \in \mathbb{N}, 0 \leq M(\alpha, \mathit{size}(n+1)) \}
\end{displaymath}
So a natural number $n$ is realised by the value
$\mathrm{natValue}(n)$ as long as we have at least
$\mathit{size}(n+1)$ potential (we add one to make the LFPL soundness
proof easier).  This gives us the ability to represent natural numbers
as a type in QTT, but in order to iterate (and construct in the case
of LFPL), we need to construct specific realisers for the
\ConsFree{} and LFPL systems.

@@

The \ConsFree{} system uses the $\MaxPoly$ resource monoid, with
the distinguished sub-monoid being those elements that are $0$ in the
size component. We enumerate the features of the \ConsFree{}
system and justify their realisability with the $\MaxPoly$ resource
monoid:
\begin{enumerate}
\item Duplication of natural numbers by $\dupNat(M)$ is realisable by
  the expression $(0,0)$, which creates a pair by copying the input
  variable twice. By the cost semantics in
  \autoref{sec:machine-model}, this takes $1$ step of computation (we
  assume that it is actually implemented via some pointer copy).  The
  resource accounting for this realiser works because the size
  component required for the output is the maximum of the size
  components of the two elements, and since $n \sqcup n = n$, we have
  enough resources to fulfil this.
\item Construction of natural numbers is not realisable. In a putative
  $\conSucc$ rule, we would need to get an additional unit of size
  resource from nowhere.
\item Iteration is realised by constructing a realising expression in
  the expression language from the given expressions for the zero and
  successor cases that uses the in-built recursion of the
  language. The proof that resources are correctly accounted for is
  carried out by induction on the natural being iterated over. For
  $n$, we require potential
  $\mathit{scale}(n,\mathit{acct}(4) \rplus \gamma_{\mathit{succ}})
  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$, where
  $\gamma_{\mathit{succ}}$ and $\gamma_{\mathit{zero}}$ are the
  potentials required by the successor and zero cases respectively. By
  Property (2) of iteration resource monoids, above, we know that
  $\mathit{raise}(\mathit{acct}(4) \rplus \gamma_{\mathit{succ}})
  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$ always
  dominates this requirement when paired with the potential
  $\mathit{size}(n)$ from the input. Therefore, this latter
  expression, plus some administrative set up costs, is the required
  potential for the whole iterator.
\end{enumerate}
Together, we have a soundness result for the \ConsFree{} system,
that ensures that every term in the $\sigma = 1$ fragment is
realisable by a \emph{correct} program that terminates in polynomial
time for all inputs:
\begin{theorem}[Soundness for the \ConsFree{} System]
  \label{thm:cons-free-soundness}
  If we have a term $n \stackrel1: \tyNat \vdash M \stackrel1: T(n)$ then there
  exists a realising expression $E$ and polynomial $p$ such that for
  all $n \in \mathbb{N}$, there exists $v \in \ValSet$ and
  $k \in \mathbb{N}$ such that
  $E, [\mathrm{natValue}(n)] \Downarrow_k v$, $k \leq p(n)$ and $v$ is
  a realising value for
  $\llbracket M \rrbracket(n) \in \llbracket T \rrbracket(n)$.
\end{theorem}

\paragraph{Agda Formalisation} The realisability of the
\ConsFree{} system iterator and the soundness property of the
whole system are formalised in the Agda modules \AgdaModule{ConsFree}
and \AgdaModule{ConsFree.Iterator}. The soundness theorem is a
combination of this and the QTT model sketched in
\autoref{sec:qtt-models}.

@@

The LFPL system uses the $\PlusPoly$ resource monoid, with the
distinguished sub-monoid again being those elements that are $0$ in
the size component. With this resource monoid, the capabilities
offered at the QTT level are altered:
\begin{enumerate}
\item We can no longer duplicate natural numbers, because
  $\tyNat \otimes \tyNat$ requires twice as much size resource as
  $\tyNat$, due to the combining operation on size potentials being
  addition.
\item We define the realisability specification for diamonds
  $\Diamond \in L(1)$ to be
  $\Diamond = \{ (*, \alpha, *) \mid 0 \leq M(\alpha,
  \mathit{size}(1)) \}$.  Thus, a diamond represents at least one unit
  of size resource, matching the intuitive explanation given in
  \autoref{sec:lfpl-intro}.
\item With this definition of realisability for $\Diamond$s, it is
  possible to realise the $\conZero$ and $\conSucc$ constructors for
  natural numbers. By the additive combination of size resources we
  get $1$ from the diamond and $n+1$ from the predecessor to total
  $n+2$ for a new number. Note that, even if we add a $\Diamond$ type
  to the \ConsFree{} system, it would still not be possible to
  realise the constructors, because we would only have
  $1 \sqcup (n+1) = n+1$ size resource for the output.
\item The construction of the recursor follows a very similar proof to
  the realisability of \ConsFree{} iterator, up to some
  additional work to make sure that the dummy $*$ values representing
  the diamond components end up in the right places. This additional
  work is revealed in the required potential for the LFPL iterator
  being
  $\mathit{raise}(\mathit{acct}(8) \rplus \gamma_{\mathit{succ}})
  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$, so
  slightly higher in the successor case.
\end{enumerate}

Soundness for the LFPL system is similar to the \ConsFree{}
system, except for a $+1$ to the input to the polynomial, to account
for the fact that we cost one size unit for the $\conZero$
constructor.
\begin{theorem}[Soundness for the LFPL-style System]
  \label{thm:lfpl-soundness}
  If we have a term $n \stackrel1: \tyNat \vdash M \stackrel1: T(n)$ then there
  exists a realising expression $E$ and polynomial $p$ such that for
  all $n \in \mathbb{N}$, there exists $v \in \ValSet$ and
  $k \in \mathbb{N}$ such that
  $E, [\mathrm{natValue}(n)] \Downarrow_k v$, $k \leq p(n+1)$ and $v$
  is a realising value for
  $\llbracket M \rrbracket(n) \in \llbracket T \rrbracket(n)$.
\end{theorem}

The proofs of well-accounted realisability for the LFPL iterator, and
the \ConsFree{} iterator, could be adapted to any other
inductively defined type that is finitely branching. This is not
immediately necessary, as evidenced by the construction of other
datatypes in \autoref{sec:prog-datatypes}. Nevertheless, native tree
type where the iterability is proportional to the total number of
nodes would be useful.

@@

We have presented two extensions of Quantitative Type Theory that
soundly and completely capture polynomial time. This allows for an
expressive combination of verification and complexity constrained
computation, including characterisations of the classes P, NP, and
BPP. We now discuss related work, and take a look at where the
combination of polytime and dependency could take us.

@@

\paragraph{Implicit Computational Complexity with Linear Types}
Implicit Computational Complexity \cite{Lago11} is a vast field, so we
only survey closely related works. We have already mentioned the
Bounded Linear Logic \cite{bll92}, Soft Affine Logic \cite{Lafont04},
Light Linear Logic \cite{lll98} and LFPL \cite{hofmann99lfpl} systems,
which all use linear typing to implicitly capture polynomial
time. \citet{Jones01} characterises polynomial time using first-order
functional programs without constructors. Thus it shares a method with
our \ConsFree{} system, but we use linear typing to permit
controlled use of higher-order functions. Other approaches to
polynomial time use stratification or information flow tracking to
ensure that the outputs of iteration may not be used unrestrictedly to
drive further iteration. For example, \cite{BellantoniC92} and
\cite{HainryP23}. Below polynomial time, systems have be devised to
capture LOGSPACE \cite{LagoS16}. Above polynomial time, systems such
as Elementary Affine Logic (EAL) capture all Elementary-time functions
\cite{CoppolaM01}.

@@

\paragraph{Other Complexity Classes}
We have been able to characterise the classes NP and BPP in terms of
our underlying characterisation of P (\autoref{sec:classes}). It seems
straighforward to extend this to related classes like coNP, RP,
etc. It also seems feasible to adapt the techniques presented here to
other complexity classes such as LOGSPACE and ELEMENTARY, given the
simply typed linear systems mentioned above. Complexity classes based
on circuits may be more challenging, but we do now have a way to
characterise circuits that are generatable in polynomial
time.
